---
title: "GWAS SurvivR Vignette"
author:
- name: Abbas Rizvi
  affiliation: The Ohio State University, Columbus, OH
- name: Ezgi Karaesmen
  affiliation: The Ohio State University, Columbus, OH
- name: Martin Morgan
  affiliation: Roswell Park Cancer Institute, Buffalo, NY    
- name: Lara Sucheston-Campbell  
  affiliation: The Ohio State University, Columbus, OH
package: GWAS SurvivR
date: "October 31, 2017"
output:
        BiocStyle::html_document2
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}          
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This package provides a workflow for how to use imputed data GWAS data from either IMPUTE2 or the Michigan server and conduct survival analysis. 

# Getting started

# IMPUTE2 data
## Convert IMPUTE2 data to GDS

## Read in files as SummarizedExperiment

## Add covariates

## Run survival

# Compressed VCF files

## Run survival

# New implementation of coxph

# Session info
Here is the output of `sessionInfo()` on the system that this document was compiled:
```{r, echo=F}
sessionInfo()
```
